---
title: "Biostat 203B Homework 4"
subtitle: "Due Mar 9 @ 11:59PM"
author: "Jiayi Guo 206537584"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

Display machine information:
```{r}
sessionInfo()
```
Display my machine memory.
```{r}
memuse::Sys.meminfo()
```

Load database libraries and the tidyverse frontend:
```{r}
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
```

## Q1. Compile the ICU cohort in HW3 from the Google BigQuery database 

Below is an outline of steps. In this homework, we exclusively work with the BigQuery database and should not use any MIMIC data files stored on our local computer. Transform data as much as possible in BigQuery database and `collect()` the tibble **only at the end of Q1.7**.

### Q1.1 Connect to BigQuery

Authenticate with BigQuery using the service account token. Please place the service account token (shared via BruinLearn) in the working directory (same folder as your qmd file). Do **not** ever add this token to your Git repository. If you do so, you will lose 50 points.
```{r}
# path to the service account token 
satoken <- "biostat-203b-2025-winter-4e58ec6e5579.json"
# BigQuery authentication using service account
bq_auth(path = satoken)
```
Connect to BigQuery database `mimiciv_3_1` in GCP (Google Cloud Platform), using the project billing account `biostat-203b-2025-winter`.
```{r}
# connect to the BigQuery database `biostat-203b-2025-mimiciv_3_1`
con_bq <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2025-winter",
    dataset = "mimiciv_3_1",
    billing = "biostat-203b-2025-winter"
)
con_bq
```

List all tables in the `mimiciv_3_1` database.
```{r}
dbListTables(con_bq)
```
```{r}
d_items_tble <- tbl(con_bq, "d_items")
admissions_tble <- tbl(con_bq, "admissions")
patients_tble <- tbl(con_bq, "patients")
d_labitems_tble <- tbl(con_bq, "d_labitems")
```
### Q1.2 `icustays` data

Connect to the `icustays` table.
```{r}
# full ICU stays table
icustays_tble <- tbl(con_bq, "icustays") |>
  arrange(subject_id, hadm_id, stay_id) |> 
  # show_query() |>
  print(width = Inf)
```


### Q1.3 `admissions` data

Connect to the `admissions` table.
```{r}
admissions_tble <- tbl(con_bq, "admissions") |>
  arrange(subject_id, hadm_id) |>
  print(width = Inf)
  
```

### Q1.4 `patients` data

Connect to the `patients` table.
```{r}
patients_tble <- tbl(con_bq, "patients") |>
  arrange(subject_id) |>
  print()
```

### Q1.5 `labevents` data

Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW3. Only keep the last lab measurements (by `storetime`) before the ICU stay and pivot lab items to become variables/columns. Write all steps in _one_ chain of pipes.

**steps**
Get the labevents
```{r}
labevents_tble <- tbl(con_bq, "labevents") |>
  semi_join(icustays_tble, by = "hadm_id") |>
  filter(itemid %in% c(50912, 50971, 50983, 50902, 
                       50882, 51221, 51301, 50931)) |>
  left_join(d_labitems_tble, by = "itemid") |>
  select(-subject_id) |>
  left_join(icustays_tble, by = "hadm_id") |>
  select(subject_id, stay_id, intime, itemid, storetime, valuenum, label) |>
  filter(storetime < intime) |>
  # Slice() isn’t supported in SQL so I use distince instead 
  arrange(stay_id, itemid, desc(storetime)) |>
  distinct(stay_id, itemid, .keep_all = TRUE) |>
  select(subject_id, stay_id, valuenum, label) |>
  pivot_wider(names_from = label, values_from = valuenum) |>
  print(width = Inf)
  
  
```

**Unresolved**
hadm_id has lots of NA value, should I do something to this?
lack lots of valuenum, is that normal?
### Q1.6 `chartevents` data

Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW3. Only keep the first chart events (by `storetime`) during ICU stay and pivot chart events to become variables/columns. Write all steps in _one_ chain of pipes. Similary to HW3, if a vital has multiple measurements at the first `storetime`, average them.
```{r}

```

```{r}
chartevents_tble <- tbl(con_bq, "chartevents") |>
  # Filter subjects appearing in icustay
  semi_join(icustays_tble, by = "stay_id") |>
  filter(itemid %in% c(220045, 220179, 220180, 223761, 220210)) |>
  left_join(d_items_tble, by = "itemid") |>
  select(-subject_id) |>
  select(-hadm_id) |>
  # Get intime and stay_id
  left_join(icustays_tble, by = "stay_id") |>
  # Divide every icu stay and item
  group_by(subject_id, stay_id, itemid) |>
  # Keep first measurement during icu stay
  filter(storetime >= intime) |>
  # Keep the smallest storetime
  filter(storetime == min(storetime)) |>
  # If there is only one measurement, the mean value is itself
  mutate(avg_value = mean(valuenum)) |>
  arrange(storetime) |>
  distinct(itemid, .keep_all = TRUE) |>
  ungroup() |>
  select(c("subject_id", "stay_id", "label", "avg_value")) |>
  pivot_wider(names_from = label, values_from = avg_value) |>
  arrange(stay_id) |>
  print(width = Inf)
```

### Q1.7 Put things together

This step is similar to Q7 of HW3. Using _one_ chain of pipes `|>` to perform following data wrangling steps: (i) start with the `icustays_tble`, (ii) merge in admissions and patients tables, (iii) keep adults only (age at ICU intime >= 18), (iv) merge in the labevents and chartevents tables, (v) `collect` the tibble, (vi) sort `subject_id`, `hadm_id`, `stay_id` and `print(width = Inf)`.

```{r}
mimic_icu_cohort <- icustays_tble |>
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) |>
  left_join(patients_tble, by = c("subject_id")) |>
  left_join(labevents_tble, by = c("subject_id", "stay_id")) |>
  left_join(chartevents_tble, by = c("subject_id", "stay_id")) |>
  # compute age at intime
  mutate(age_intime = anchor_age + year(intime) - anchor_year) %>%
  # keep only patients aged over 18 at intime
  filter(age_intime > 18) %>%
  collect() |>
  arrange("subject_id", "hadm_id", "stay_id") |>
  print(width = Inf)
```

### Q1.8 Preprocessing

Perform the following preprocessing steps. (i) Lump infrequent levels into "Other" level for `first_careunit`, `last_careunit`, `admission_type`, `admission_location`, and `discharge_location`. (ii) Collapse the levels of `race` into `ASIAN`, `BLACK`, `HISPANIC`, `WHITE`, and `Other`. (iii) Create a new variable `los_long` that is `TRUE` when `los` is greater than or equal to 2 days. (iv) Summarize the data using `tbl_summary()`, stratified by `los_long`. Hint: `fct_lump_n` and `fct_collapse` from the `forcats` package are useful.
**step(i)**
```{r}
table(mimic_icu_cohort$first_careunit) 
mimic_icu_cohort$first_careunit <- fct_lump(mimic_icu_cohort$first_careunit,
                                            n = 7)
table(mimic_icu_cohort$last_careunit)
mimic_icu_cohort$last_careunit <- fct_lump(mimic_icu_cohort$last_careunit,
                                            n = 7)
table(mimic_icu_cohort$admission_type)
mimic_icu_cohort$admission_type <- fct_lump(mimic_icu_cohort$admission_type,
                                            n = 4)
table(mimic_icu_cohort$admission_location)
mimic_icu_cohort$admission_location <- fct_lump(
  mimic_icu_cohort$admission_location, 
  n = 4)
table(mimic_icu_cohort$discharge_location)
mimic_icu_cohort$discharge_location <- fct_lump(
  mimic_icu_cohort$discharge_location,
  n = 4)
```
**step(ii)**
```{r}
table(mimic_icu_cohort$race)
mimic_icu_cohort <- mimic_icu_cohort |>
  mutate(
    race_total = case_when(
    str_detect(race, regex("ASIAN", ignore_case = TRUE)) ~ "ASIAN",
    str_detect(race, regex("BLACK", ignore_case = TRUE)) ~ "BLACK",
    str_detect(race, regex("HISPANIC", ignore_case = TRUE)) ~ "HISPANIC",
    str_detect(race, regex("WHITE", ignore_case = TRUE)) ~ "WHITE",
    TRUE ~ "Other"
    )
  ) |>
  select(-race) |>
  rename(race = race_total)
table(mimic_icu_cohort$race)
```
**(iii)**
```{r}
mimic_icu_cohort <- mimic_icu_cohort |>
  mutate(
    los_long = if_else(los >= 2, TRUE, FALSE)
  ) 
```
**(iv)**
```{r}
tbl_summary(mimic_icu_cohort, 
            by = los_long,
            include = c(
              "first_careunit", 
              "last_careunit", 
              "admission_type", 
              "admission_location", 
              "discharge_location",
              "race"
            ))
```



Hint: Below is a numerical summary of my tibble after preprocessing:

<iframe width=95% height="500" src="./mimic_icu_cohort_gtsummary.html"></iframe>

### Q1.9 Save the final tibble

Save the final tibble to an R data file `mimic_icu_cohort.rds` in the `mimiciv_shiny` folder.
```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
# save the final tibble
mimic_icu_cohort |>
  write_rds("mimiciv_shiny/mimic_icu_cohort.rds", compress = "gz")
```
Close database connection and clear workspace.
```{r}
if (exists("con_bq")) {
  dbDisconnect(con_bq)
}
rm(list = ls())
```
Although it is not a good practice to add big data files to Git, for grading purpose, please add `mimic_icu_cohort.rds` to your Git repository.

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should contain at least two tabs. One tab provides easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort, using the `mimic_icu_cohort.rds` you curated in Q1. The other tab allows user to choose a specific patient in the cohort and display the patient's ADT and ICU stay information as we did in Q1 of HW3, by dynamically retrieving the patient's ADT and ICU stay information from BigQuery database. Again, do **not** ever add the BigQuery token to your Git repository. If you do so, you will lose 50 points.
```{r}
subject_id <- 10063848
ggplot() +
  geom_segment(
    data = ADT10063848 |>
      filter(careunit != "UNKNOWN"),
    aes(
      x = intime_date,
      xend = outtime_date,
      y = "ADT",
      yend = "ADT",
      color = careunit,
      linewidth = is_icu
    )
  ) +
  geom_point(
  data = labs10063848,
  aes(x = chartdate, y = "Lab"),
  shape = 4, size = 3
  ) +
  geom_point(
    data = procedures10063848,
    aes(x = chartdate, y = "Procedure", 
        shape = factor(shape), color = procedure_type),
    size = 3
  ) +
  scale_shape_manual(values = c(16, 17, 15, 19)) +
  labs(title = "Patient 10063848, F, 75 years old, white",
       subtitle = diagnoses_text,
       x = "Date",
       y = "",
       shape = "Procedure Type",
       color = "Procedure Type") +
  theme_minimal()

```




